---
title: "Mapping Environmental Injustice in LA County"
author: Joshua Ferrer-Lozano
date: 11/10/2025
format: html
code-fold: False
knitr: 
  opts_chunk: 
    warning: false
    message: false
---
# Reading in Data

```{r,warning=FALSE, message=FALSE}
#Importing libraries
suppressPackageStartupMessages({
library(sf)
library(tmap)
library(tidyverse)
library(here)
library(gt)
library(dplyr)
library(kableExtra)
library(knitr)
sf::sf_use_s2(FALSE) # avoid S2 indexing error
})
```

```{r,warning=FALSE, message=FALSE}
#Reading in data

#Initializing variables for our data
ejscreen <- st_read(here("data", "ejscreen", "ejscreen_2023_statepct.gdb"), quiet = TRUE)

#Read in HOLC data
holc <- st_read(here("data", "mapping-inequality","mapping-inequality-los-angeles.json"), quiet = TRUE)

#Read in bird data
birds <- st_read(here("data", "gbif-birds-LA"), quiet = TRUE)

#Ensuring data sets have the same CRS
ejscreen <- st_transform(ejscreen, st_crs(holc))
#Unit check 
st_crs(holc, parameters = TRUE)$units_gdal
```


```{r,warning=FALSE, message=FALSE}

#Creating plot of HOLC polygons overlaying LA county
tmap_mode("plot")

tm_shape(holc) +
  tm_basemap(server = "OpenStreetMap") +
  tm_polygons(col = "grade",
#Using custom colors to better match original color scheme
              palette = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B"),
              title = "HOLC Grade",
              labels = c('A'='Best (Green)','B'='Still Desireable (Blue)','C' = 'Declining (Yellow)', 'D'='Hazardous (Red)'),
              alpha = 0.6,
              border.col = "black",
              border.lwd = 0.5,
              ) +
  tm_layout(
    main.title = "Historical Redlining in Los Angeles",
    legend.outside = TRUE,
    frame = FALSE
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))

```

```{r,warning=FALSE, message=FALSE}

#Perform spatial join to assign HOLC grades to block groups
joined_ej_holc <- st_join(ejscreen, holc, join = st_intersects)

#Replace NA grades with "Ungraded"
joined_ej_holc <- joined_ej_holc %>%
  mutate(grade = ifelse(is.na(grade), "Ungraded", grade))

#Create summary table
summary_table <- joined_ej_holc %>%
  st_drop_geometry() %>%
  na.omit() %>%
  group_by(grade) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percent = round(count / sum(count) * 100, 1))

#Display the table
summary_table %>%
  kable(caption = "Percentage of Census Block Groups Within HOLC Grade",
        col.names = c("HOLC Grade", "Count", "Percent (%)")) %>%
  kable_styling(full_width = FALSE)

```

```{r}
#Group and summariZE EJScreen variables
ej_summary <- joined_ej_holc %>%
  group_by(grade) %>%
  summarize(
#Get means of our target variables %low income, percentile life expectancy and particulate matter
    mean_low_income = mean(LOWINCOME, na.rm = TRUE),
    mean_pm25 = mean(PM25, na.rm = TRUE),
    mean_life_expectancy = mean(LIFEEXPPCT, na.rm = TRUE)
  )
```

```{r,warning=FALSE, message=FALSE}
#Plot % percent low income
ggplot(ej_summary, aes(x = grade, y = mean_low_income, fill = grade)) +
  geom_col() +
  labs(title = "% Low Income by HOLC Grade", x = "HOLC Grade", y = "Mean % Low Income") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

```{r,warning=FALSE, message=FALSE}
#Plot PM2.5
ggplot(ej_summary, aes(x = grade, y = mean_pm25, fill = grade)) +
  geom_col() +
  labs(title = "PM2.5 Percentile by HOLC Grade", x = "HOLC Grade", y = "Mean Percentile") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

```{r,warning=FALSE, message=FALSE}
#Plotting life expectancy
ggplot(ej_summary, aes(x = grade, y = mean_life_expectancy, fill = grade)) +
  geom_col() +
  labs(title = "Low Life Expectancy Percentile by HOLC Grade", x = "HOLC Grade", y = "Mean Percentile") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()
```
## Interpret the patterns you observe in your results

What I observed through this assignment was not atypical. Residents in lower HOLC graded areas tend to be of the lower socioeconomic status, and are prone to higher levels of environmental injustices (zoning and lack of green space) and social obstacles. In this example, higher PM2.5 levels and "coincidentally" higher instance of lower live expectancy. 

Moreover, it is startling to know there are a lot of NA values. This illustrates that there is far more that we do not know about the sample/population in terms of the affects of historic HOLC ratings.This may be due to historical urban sprawl where the expansion of LA.

## Discuss potential relationships between historical redlining grades and current environmental/socioeconomic conditions

Based on the observations, HOLC rated areas tend to have an inverse relationship between the rating and the number of lower income persons. With graded A areas having the least and D having the higher percentage of lower income persons.Fast forward, this "established" methodology can and has allowed redlining policies to: keep lower income people in certain areas, hinder investment in resources, create a lack of green space/recreation centers, limit education/job opportunities, unfair zoning (refineries,factories freeways).

# Part 2:Legacy of redlining in biodiversity observations

```{r,warning=FALSE, message=FALSE}
#Transforming the 
birds <- st_transform(birds, st_crs(holc))

#Spatial join: assign HOLC grade to each bird observation
bird_joined <- st_join(birds, holc)

#Drop geometry and summarize proportions based on HOLC grades
bird_summary <- bird_joined %>%
  st_drop_geometry() %>%
  na.omit() %>%
  group_by(grade) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(percent = round(count / sum(count) * 100, 1))

#Plot bird observations
ggplot(bird_summary, aes(x = grade, y = percent, fill = grade)) +
  geom_col() +
  labs(title = "Bird Observations by HOLC Grade in LA", x = "HOLC Grade", y = "Percent of Observations") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

## Our results don’t match the findings from Ellis-Soto et al. 2023! Read the abstract of the study. Why might we have obtained different results in our analysis? What did the paper consider that we did not? 

Our analysis differed because of a few things: We focused on Los Angeles County whereas they focused on a country-wide analysis. They adjusted their approach also with confounding variables such as vegetation, open-space, population density and climate. They also used a mixed effects model, whereas we used sampling density alone. Lastly Ellis-Soto mentions survey completeness.





