---
title: "HW2"
format: html
code-fold: False
knitr: 
  opts_chunk: 
    warning: false
    message: false
---

```{r}
#importing libraries
library(sf)
library(tmap)
library(tidyverse)
library(here)
library(gt)
library(dplyr)
```

```{r}
#reading in data

# initializing variables for our data
ejscreen <- st_read(here("data", "ejscreen", "ejscreen_2023_statepct.gdb"))

holc <- st_read(here("data", "mapping-inequality","mapping-inequality-los-angeles.json"))

birds <- st_read(here("data", "gbif-birds-LA"))

# ensuring datasets have the same CRS
ejscreen <- st_transform(ejscreen, st_crs(holc))
```


```{r}

# creating plot of HOLC polygons overlaying LA county
tmap_mode("plot")

tm_shape(holc) +
  tm_basemap(server = "OpenStreetMap") +
  tm_polygons(col = "grade",
              #using custom colots to better match original color scheme
              palette = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B"),
              title = "HOLC Grade",
              alpha = 0.6,
              border.col = "black",
              border.lwd = 0.5) +
  tm_layout(
    main.title = "Historical Redlining in Los Angeles",
    legend.outside = TRUE,
    frame = FALSE
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("right", "top"))

```

```{r}
#had error about indexes and S2, needed to set to FALSE

sf::sf_use_s2(FALSE)
#assigning HOLC grades to each block group
joined_ej_holc <- st_join(ejscreen, holc)

# dropping geometries for the summary
joined_ej_holc <- st_drop_geometry(joined_ej_holc)

# counting block groups by HOLC grade
summary_table <- joined_ej_holc %>%
  mutate(grade = ifelse(is.na(grade), "Ungraded", grade)) %>%
  count(grade) %>%
  mutate(percent = round(100 * n / sum(n), 1))

# displaying a table with the r package gt
summary_table %>%
  gt() %>%
  tab_header(title = "Block Groups by HOLC Grade") %>%
  fmt_number(columns = vars(percent), decimals = 1)

```
```{r}
# grouping and summarizing EJScreen variables
ej_summary <- joined_ej_holc %>%
  group_by(grade) %>%
  summarize(
    #getting means of our target variables %low income, percentile life expectancy and particulate matter
      mean_low_income = mean(LOWINCOME, na.rm = TRUE),
    mean_pm25 = mean(PM25, na.rm = TRUE),
    mean_life_expectancy = mean(LIFEEXPPCT, na.rm = TRUE)
  )
```

```{r}

# plotting % percent low income
ggplot(ej_summary, aes(x = grade, y = mean_low_income, fill = grade)) +
  geom_col() +
  labs(title = "% Low Income by HOLC Grade", x = "HOLC Grade", y = "Mean % Low Income") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

```{r}
# plotting PM2.5
ggplot(ej_summary, aes(x = grade, y = mean_pm25, fill = grade)) +
  geom_col() +
  labs(title = "PM2.5 Percentile by HOLC Grade", x = "HOLC Grade", y = "Mean Percentile") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

```{r}
# plotting life expectancy
ggplot(ej_summary, aes(x = grade, y = mean_life_expectancy, fill = grade)) +
  geom_col() +
  labs(title = "Low Life Expectancy Percentile by HOLC Grade", x = "HOLC Grade", y = "Mean Percentile") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()
```
## Interpret the patterns you observe in your results

What I observed through this assignment was not atypical. Residents in lower HOLC graded areas tend to be of the lower socioeconomic status, and are prone to higher levels of environmental injustices (zoning and lack of green space) and social obstacles. In this example, higher PM2.5 levels and "coincidentally" higher instance of lower live expectancy. 

Moreover, it is startling to know there are a lot of NA values. This illustrates that there is far more that we do not know about the sample/population in terms of the affects of historic HOLC ratings.

## Discuss potential relationships between historical redlining grades and current environmental/socioeconomic conditions

Based on the observations, HOLC rated areas tend to have an inverse relationship between the rating and the number of lower income persons. With graded A areas having the least and D having the higher percentage of lower income persons.Fast forward, this "established" methodology can and has allowed redlining policies to: keep lower income people in certain areas, hinder investment in resources, create a lack of green space/recreation centers, limit education/job opportunities, unfair zoning (refineries,factories freeways).

# Part 2:Legacy of redlining in biodiversity observations

```{r}
# transforming the 
birds <- st_transform(birds, st_crs(holc))

# Spatial join: assign HOLC grade to each bird observation
bird_joined <- st_join(birds, holc)

# Drop geometry and summarize
bird_summary <- bird_joined %>%
  st_drop_geometry() %>%
  count(grade) %>%
  mutate(percent = round(100 * n / sum(n), 1))

# Plot bird observations
ggplot(bird_summary, aes(x = grade, y = percent, fill = grade)) +
  geom_col() +
  labs(title = "Bird Observations by HOLC Grade", x = "HOLC Grade", y = "Percent of Observations") +
  scale_fill_manual(values = c("A" = '#3C9D32',"B" = '#3A7CA5', "C" = "#F4D35E", "D" = "#D1495B")) +
  theme_minimal()

```

## Our results don’t match the findings from Ellis-Soto et al. 2023! Read the abstract of the study. Why might we have obtained different results in our analysis? What did the paper consider that we did not? 

Our analysis differed because of a few things:
- We focused on Los Angeles but they did a country-wide analysis
- They adjusted their approach woth confounding gvariables such as vegetation, open space, population desnity and climate
- They used a mixed effects model
- We used sampling density alone but Ellis-Soto mentioned survey completeness.


